### 2023-09-27 main.tex
clean

\begingroup
% \centering
\begin{center}
 \begingroup\LARGE
\bf Multispecies stirring process
\par\endgroup
 \vspace{3.5em}
 \begingroup\large \bf
% {Rouven Frassek}
 \par\endgroup
\vspace{2em}

@@

The way one can define a multi-type (or multi-species) interacting particle system is not unique. In the literature there are many. Two of them are the following. On one hand, the \textit{multi-layer models}, in which the process is constructed on a many layer graph where each graph correspond to a type and perform a kind of dynamic (independent jump, exclusion/inclusion walkers) and the layers communicate each other with some "switching probabilities". The state space of the Markov process is the Cartesian product of the state spaces of each layer, i.e. of the two species. For example see \cite{floreani2022switching,redig2022ergodic}. On the other hand, the \textit{single layer models}. Here, there is just one layer, where particles of each type can pile up, in a way that the exclusion dynamic concerns many species together. In other words, now the species can interact "directly", not just by switching layer. Here the state space contains all the species and it is not product anymore. Some examples of such a dynamic can be found in \cite{casini2022uphill,vanicat2017exact,zhou2021orthogonal}. \\

@@

The case of $\twoj=1$ distinguishes itself from other values of $\twoj$ as in this case the model can be mapped to an integrable higher rank $gl(n)$ Heisenberg XXX spin chain with open (non-diagonal) boundaries. Furthermore the matrix product ansatz \cite{derrida1993exact}  is applicable, see \cite{Crampe:2014aoa,vanicat2017exact}.

@@

See Figure \ref{fig:1} and  Figure \ref{fig:2} for a visualisation of the dyamics. 
\begin{figure}

@@

\begin{figure}

@@

{\color{blue}check also Appendix A in

@@

The boundary generator $\mathcal{L}_{right}^{rd}$ is defined analugosly, but the $\alpha$'s are replaced by $\beta$'s and the boundary densities are considered at site $L+1$. \\ \\

@@

	\Lambda^{x}_{rev}\sim \text{Multinomial}\left(\twoj,\rho_{1},\ldots,\rho_{N}\right)

@@

	\rho=\rho_{a}=\frac{\alpha}{\alpha_{N}+(N-1)\alpha}\quad \forall a\in \{1,\ldots,N-1\}\qquad \text{and}\qquad \rho_{N}=\frac{\alpha_{N}}{\alpha_{N}+(N-1)\alpha}

@@

	\Lambda_{rev}^{x}(n^{x})=\frac{\nu!}{\prod_{a=1}^{N}n_{a}^{x}!}\rho^{\sum_{a=1}^{N-1}n_{a}^{x}}\rho_{N}^{n_{N}^{x}}
This can be proved just by imposing the detailed balance conditions for the edge $x,x+1$ and for the boundary generators. For fixed site $x\in \{1,\ldots,L\}$ we only prove the detailed balance for  $\mathcal{L}_{x,x+1}^{c}$ with $c\in\{1,\ldots,N-2\}$ since for the other generators  $\mathcal{L}_{x,x+1}^{m}$, $\mathcal{L}_{left}^{rd}$ and $\mathcal{L}_{right}^{rd}$ the computation is similar. Consider the transition $\bm{n}\to \bm{n}-\delta_{a}^{x}+\delta_{g_{c}(b)}^{x}+\delta_{g_{c}(b)}^{x+1}-\delta_{b}^{x+1}$ for arbitrary $a,b\in\{1,\ldots,N-1\}$ and $c\in\{1,\ldots,N-2\}$; the detailed balance reads

@@

	\rho_{a}\rho_{b}=\rho_{g_{c}(a)}\rho_{g_{c}(b)}
By the arbitrariness of $a,b,$ it is true if and only if $\rho_{a}=\rho$ $\forall a\in \{1,\ldots,N-1\}$.\\
For the boundary transitions we have that, analogously to what happens for the multispecies stirring process, the reservoirs parameters must be such that $\beta_{a}=\alpha_{a}$.

@@

The dual  process $(\bm{\xi}(t))_{t\geq 0}$ still takes values in the enlarged state space \eqref{stateSpaceDUAL-Chain}

@@

We denote the configurations $\bm{\xi} \in \widetilde\Omega$ 

@@

The dual process is defined by the dual generator

@@

It is enough of showing 

@@

 We compute the commutator between $\mathcal{H}_{x,x+1}^{c}$ and $\sum_{d=1}^{N-1}(E_{dN}^{x})\otimes \sum_{k=1}^{N-1}(E_{bN}^{x})$. Using \eqref{eq:comgl}, the bilinearity and associativity of the Kronecker product and the bilinearity of the brackets we obtain 

@@

	\sum_{d,k=1}^{N-1}\sum_{a,b=1}^{N}\left\{\left((E_{da}^{x}\delta_{g_{c}(b)N}-E_{g_{c}(b)N}^{x}\delta_{ad})-(E_{da}^{x+1}\delta_{aN}-E_{aN}^{x+1}\delta_{da})\right)\right. \\ \otimes&\left. \left((E_{kb}^{y}\delta_{g_{c}(a)N}-E_{g_{c}N}^{x+1}\delta_{kb})-(E_{kb}^{x+1}\delta_{bN}-E_{bN}^{x+1}\delta_{kb})\right)\right\}
	\sum_{d,k=1}^{N-1}\sum_{a,b=1}^{N}\left\{\left(E_{da}^{x}-E_{g_{c}(b)N}^{x}-E_{da}^{x}+E_{aN}^{x}\right)\otimes\left(E_{kb}^{x+1}-E_{g_{c}(a)N}^{x+1}-E_{kb}^{y}+E_{bN}^{x+1}\right)\right\}

@@

where in the last equality we used the invertibility of the map $g_{c}(\cdot)$. \\

@@

We compute the commutator between $\mathcal{H}_{x,x+1}^{m}$ and $\sum_{d=1}^{N-1}(E_{dN}^{x})\otimes \sum_{k=1}^{N-1}E_{bN}^{x}$. The prof is similar and gives

@@

\section{Conclusion}

@@

	\paragraph{Proof of formula \eqref{steadyStateH}} As a consequence of \eqref{S1-Inverse}, we only need to show how the transformation $S_{1}^{-1}$ acts on the vector $|\Psi^{'}\rangle$. Here is more convenient to consider the matrix $\mathcal{S}_{1}^{-1}$ defined in \eqref{S2Gen}, acting on a single site, and make explicitly the tensor product over the chain. This matrix $\mathcal{S}_{1}$ is given by
	\begin{equation}\label{S1-L1}
		\mathcal{S}_{1}^{-1}=\sum_{\tau_{1}=1}^{N}e_{\tau_{1}}^{\tau_{1}}+(-1)\delta_{\tau_{1},N}\sum_{d_{1}=1}^{N-1}e_{1}^{d_{1}}
	\end{equation}
	Using the notation introduced in \eqref{notation1} and \eqref{notation2} we have that 
	\begin{equation}
		\begin{split}
			\left(\mathcal{S}_{1}^{-1}\right)^{\otimes L}&=\sum_{\tau_{1}=1}^{N}\ldots\sum_{\tau_{L}=1}^{N}\left\{e_{\tau_{1},\ldots,\tau_{L}}^{\tau_{1},\ldots,\tau_{L}}\right.
			\\&
			+\left.\sum_{b=1}^{L}(-1)^{b}\sum_{q_{1}=1}^{L}\sum_{q_{2}=q_{1}+1}^{L}\ldots\sum_{q_{b}=q_{b-1}+1}^{L}\left(\prod_{r=1}^{b}\delta_{s_{q_{r}},N}\right)\sum_{d_{q_{1}}=1}^{N-1}\ldots\sum_{d_{q_{r}}=1}^{N-1}e_{\eta_{1},\ldots,\eta_{L}}^{\theta_{1},\ldots,\theta_{L}}\right\} 
		\end{split}
	\end{equation}
	where we have introduced, $\forall x\in \{1,\ldots,L\}$
	\begin{equation}
		%\eta_{x}=\begin{cases}
			%	\tau_{x}\qquad &\text{if}\quad x\neq 1,\ldots,b\\
			%	d_{x}\qquad &\text{otherwise}
			%\end{cases}\qquad
			\theta_{x}(b)=\begin{cases}
				\tau_{x}\qquad &\text{if}\quad x\notin \{q_{1},\ldots,q_{b}\}\\
				1 \qquad &\text{otherwise}
			\end{cases}
		\end{equation}
		The dependence of $\theta(b)$ with respect to $b$ is usually understood for the sake of notation. 
		This is proved by induction. Indeed, for $L=1$ it is trivially true, by \eqref{S1-L1}. We assume that it is true for arbitrary $L\in \mathbb{N}$ and we prove that it holds for $L+1$. 
		\begin{align*}
			\left(\mathcal{S}_{1}^{-1}\right)^{\otimes L}\otimes \mathcal{S}_{1}^{-1}&=
			\sum_{\tau_{1}=1}^{N}\ldots\sum_{\tau_{L}=1}^{N}\left\{e_{\tau_{1},\ldots,\tau_{L}}^{\tau_{1},\ldots,\tau_{L}}\right. \\&+\left. \sum_{b=1}^{L}(-1)^{b}\sum_{q_{1}=1}^{L}\sum_{q_{2}=q_{1}+1}^{L}\ldots\sum_{q_{b}=q_{b-1}+1}^{L}\left(\prod_{r=1}^{b}\delta_{s_{q_{r}},N}\right)\sum_{d_{q_{1}}=1}^{N-1}\ldots\sum_{d_{q_{r}}=1}^{N-1}e_{\theta_{1},\ldots,\theta_{L}}^{\eta_{1},\ldots,\eta_{L}}\right\}
			\\&\otimes
			\left\{ \sum_{\tau_{1}=1}^{N}e_{\tau_{L+1}}^{\tau_{L+1}}+(-1)\delta_{\tau_{L+1},N}\sum_{d_{L+1}=1}^{N-1}e_{1}^{d_{L+1}}\right\}
			\\&=
			\sum_{\tau_{1}=1}^{N}\ldots\sum_{\tau_{L+1}=1}^{N}\left\{e_{\tau_{1},\ldots,\tau_{L+1}}^{\tau_{1},\ldots,\tau_{L+1}}\right. \\&+\left. \sum_{b=1}^{L+1}(-1)^{b}\sum_{q_{1}=1}^{L+1}\sum_{q_{2}=q_{1}+1}^{L+1}\ldots\sum_{q_{b}=q_{b-1}+1}^{L+1}\left(\prod_{r=1}^{b}\delta_{s_{q_{r}},N}\right)\sum_{d_{q_{1}}=1}^{N-1}\ldots\sum_{d_{q_{b}}=1}^{N-1}e_{\theta_{1},\ldots,\theta_{L+1}}^{\eta_{1},\ldots,\eta_{L+1}}\right\}
		\end{align*}
		\begin{comment}

@@

indeed, in each coordinate $q_{r}$ with $r=1,\ldots b$ the species $1$ is replaced by all other possible species from $2$ to $N$.  Finally, considering all possible $\tau_{1},\ldots,\tau_{L}\in \{1,\ldots,N\}$, we obtain \eqref{steadyStateH}. 

>>>>

\begingroup
% \centering
\begin{center}
 \begingroup\LARGE
\bf Multispecies stirring process: duality and exact solution
\par\endgroup
 \vspace{3.5em}
 \begingroup\large \bf
% {Rouven Frassek}
 \par\endgroup
\vspace{2em}

@@

The way one can define a multi-type (or multi-species) interacting particle system is not unique. In the literature there are many. Two of them are the following. On one hand, the \textit{multi-layer models}, in which the process is constructed on a many layer graph where each graph correspond to a type and perform a kind of dynamic (independent jump, exclusion/inclusion walkers) and the layers communicate each other with some "switching probabilities". The state space of the Markov process is the Cartesian product of the state spaces of each layer, i.e. of the two species. For example see \cite{floreani2022switching,redig2022ergodic,redig2023equilibrium}. On the other hand, the \textit{single layer models}. Here, there is just one layer, where particles of each type can pile up, in a way that the exclusion dynamic concerns many species together. In other words, now the species can interact "directly", not just by switching layer. Here the state space contains all the species and it is not product anymore. Some examples of such a dynamic can be found in \cite{casini2022uphill,vanicat2017exact,zhou2021orthogonal}. \\

@@

The case of $\twoj=1$ distinguishes itself from other values of $\twoj$ as in this case the model can be mapped to an integrable higher rank $gl(N)$ Heisenberg XXX spin chain with open (non-diagonal) boundaries. Furthermore the matrix product ansatz \cite{derrida1993exact}  is applicable, see \cite{Crampe:2014aoa,vanicat2017exact}.

@@

See Figure \ref{fig:1} and  Figure \ref{fig:2} for a visualisation of the dynamics. 
\begin{figure}[h]

@@

\begin{figure}[h]

@@

{\color{red}check also Appendix A in

@@

As a consequence, we can define the boundary values in terms of the reservoir parameters as
	\rho_{a}^{0}=\frac{\alpha_{a}\sigma_{11}-\sigma_{12}\sum_{b=1\,:\,b\neq a}^{N-1}\alpha_{b}}{(\sigma_{11}-(N-2)\sigma_{12})(\sigma_{11}+(N-2)\sigma_{12})}\qquad \rho_{N}^{0}=\frac{\alpha_{N}}{(\sigma_{11}+(N-2)\sigma_{12})}
for every $a\in\{1,\ldots,N\}$. Moreover, with this choice we have that $\sum_{a=1}^{N}\alpha_{a}=\nu(\sigma_{11}+(N-1)\sigma_{12})$ and $\sum_{a=1}^{N}\rho_{a}^{0}=\nu$. \\
The boundary generator $\mathcal{L}_{right}^{rd}$ is defined analogously, but the $\alpha$'s are replaced by $\beta$'s and the boundary values are considered at site $L+1$. \\ \\

@@

	\Lambda^{x}_{rev}\sim \text{Multinomial}\left(\twoj,p_{1},\ldots,p_{N}\right)

@@

	p=p_{a}=\frac{\alpha}{\alpha_{N}+(N-1)\alpha}\quad \forall a\in \{1,\ldots,N-1\}\qquad \text{and}\qquad p_{N}=\frac{\alpha_{N}}{\alpha_{N}+(N-1)\alpha}

@@

	\Lambda_{rev}^{x}(n^{x})=\frac{\nu!}{\prod_{a=1}^{N}n_{a}^{x}!}p^{\sum_{a=1}^{N-1}n_{a}^{x}}p^{n_{N}^{x}}
This can be proved just by imposing the detailed balance conditions for the edge $x,x+1$ and for the boundaries. For fixed site $x\in \{1,\ldots,L\}$ we only prove the detailed balance for  $\mathcal{L}_{x,x+1}^{c}$ with $c\in\{1,\ldots,N-2\}$ since for the generator $\mathcal{L}_{x,x+1}^{m}$ the computation is similar and has already been proved for $\mathcal{L}_{x,x+1}$ of the multi-species stirring. Consider the transition $\bm{n}\to \bm{n}-\delta_{a}^{x}+\delta_{g_{c}(b)}^{x}+\delta_{g_{c}(b)}^{x+1}-\delta_{b}^{x+1}$ for arbitrary $a,b\in\{1,\ldots,N-1\}$ and $c\in\{1,\ldots,N-2\}$; the detailed balance gives

@@

	p_{a}p_{b}=p_{g_{c}(a)}p_{g_{c}(b)}
By the arbitrariness of $a,b,$ it is true if and only if
\begin{equation}\label{p-equal}
	p_{a}=p\qquad \forall a\in \{1,\ldots,N-1\}
while $p_{N}=1-\sum_{a=1}^{N-1}p_{a}$\\
For the boundary generators $\mathcal{L}_{left}^{rd}$ and $\mathcal{L}_{right}^{rd}$ we have that, analogously to what happens for the multispecies stirring process, the reservoirs parameters must be such that $\beta_{a}=\alpha_{a}$ and, due to \eqref{p-equal}, we also need $\alpha_{a}=\alpha$ for all $a\in\{1,\ldots,N-1\}$. 

@@

The dual  process $(\bm{\xi}(t))_{t\geq 0}$ still takes values in the enlarged state space $\widetilde{\Omega}$ defined in \eqref{stateSpaceDUAL-Chain}

@@

%and we denote its configurations by $\bm{\xi} \in \widetilde\Omega$.

@@

The dual generator now reads

@@

but it is enough of showing 

@@

 We compute the commutator between $\mathcal{H}_{x,x+1}^{c}$ and $\sum_{d=1}^{N-1}(E_{dN}^{x})\otimes \sum_{k=1}^{N-1}(E_{bN}^{x+1})$. Using \eqref{eq:comgl}, the bilinearity and associativity of the Kronecker product and the bilinearity of the brackets we obtain 

@@

	\sum_{d,k=1}^{N-1}\sum_{a,b=1}^{N}\left\{\left((E_{da}^{x}\delta_{g_{c}(b)N}-E_{g_{c}(b)N}^{x}\delta_{ad})-(E_{da}^{x}\delta_{aN}-E_{aN}^{x}\delta_{da})\right)\right. \\ \otimes&\left. \left((E_{kb}^{x+1}\delta_{g_{c}(a)N}-E_{g_{c}N}^{x+1}\delta_{kb})-(E_{kb}^{x+1}\delta_{bN}-E_{bN}^{x+1}\delta_{kb})\right)\right\}
	\sum_{d,k=1}^{N-1}\sum_{a,b=1}^{N}\left\{\left(E_{da}^{x}-E_{g_{c}(b)N}^{x}-E_{da}^{x}+E_{aN}^{x}\right)\otimes\left(E_{kb}^{x+1}-E_{g_{c}(a)N}^{x+1}-E_{kb}^{x+1}+E_{bN}^{x+1}\right)\right\}

@@

where in the last equality we used the fact that map $g_{c}(\cdot)$ is surjective. \\

@@

Concerning the commutator between $\mathcal{H}_{x,x+1}^{m}$ and $\sum_{d=1}^{N-1}(E_{dN}^{x})\otimes \sum_{k=1}^{N-1}(E_{bN}^{x+1})$, the proof is similar and gives

@@

\section{Conclusion and outlook}
Multi-species version of non-compact chains. Extension of \cite{frassek2022exact},\cite{frassek2022integrable},\cite{frassek2020non},\cite{frassek2021duality}. \\
Use exact solution to compute large deviation rate function extending to the multi-species case results proved in \cite{derrida1998exact}, \cite{mallick2022exact}. \\
Extend the result to a-symmetric situation. See \cite{belitsky2018self}, \cite{belitsky2015quantum}, \cite{crampe2016integrable}, \cite{finn2018matrix}. \\
Extend the QISM technique to models which include reaction trying to apply the method exposed in \cite{de2021yang} for stirring process with a proper reaction term. Extend the duality to stirring models with different mechanism of transition \cite{chatterjee2023multi}.

@@

	\paragraph{Proof of formula \eqref{steadyStateH}} As a consequence of \eqref{S1-Inverse}, we only need to show how the transformation $S_{1}^{-1}$ acts on the vector $|\Psi^{'}\rangle$. Here is more convenient to consider the matrix $\mathcal{S}_{1}^{-1}$ defined in \eqref{S2Gen}, acting on a single site, and make explicitly the tensor product over the chain. This matrix $\mathcal{S}_{1}$ is given by
	\begin{equation}\label{S1-L1}
		\mathcal{S}_{1}^{-1}=\sum_{\tau_{1}=1}^{N}e_{\tau_{1}}^{\tau_{1}}+(-1)\delta_{\tau_{1},N}\sum_{d_{1}=1}^{N-1}e_{1}^{d_{1}}
	\end{equation}
	Using the notation introduced in \eqref{notation1} and \eqref{notation2} we have that 
	\begin{equation}
		\begin{split}
			\left(\mathcal{S}_{1}^{-1}\right)^{\otimes L}&=\sum_{\tau_{1}=1}^{N}\ldots\sum_{\tau_{L}=1}^{N}\left\{e_{\tau_{1},\ldots,\tau_{L}}^{\tau_{1},\ldots,\tau_{L}}\right.
			\\&
			+\left.\sum_{b=1}^{L}(-1)^{b}\sum_{q_{1}=1}^{L}\sum_{q_{2}=q_{1}+1}^{L}\ldots\sum_{q_{b}=q_{b-1}+1}^{L}\left(\prod_{r=1}^{b}\delta_{s_{q_{r}},N}\right)\sum_{d_{q_{1}}=1}^{N-1}\ldots\sum_{d_{q_{r}}=1}^{N-1}e_{\eta_{1},\ldots,\eta_{L}}^{\theta_{1},\ldots,\theta_{L}}\right\} 
		\end{split}
	\end{equation}
	where we have introduced, $\forall x\in \{1,\ldots,L\}$
	\begin{equation}
		%\eta_{x}=\begin{cases}
			%	\tau_{x}\qquad &\text{if}\quad x\neq 1,\ldots,b\\
			%	d_{x}\qquad &\text{otherwise}
			%\end{cases}\qquad
			\theta_{x}(b)=\begin{cases}
				\tau_{x}\qquad &\text{if}\quad x\notin \{q_{1},\ldots,q_{b}\}\\
				N \qquad &\text{otherwise}
			\end{cases}
		\end{equation}
		The dependence of $\theta(b)$ with respect to $b$ is usually understood for the sake of notation. 
		This is proved by induction. Indeed, for $L=1$ it is trivially true, by \eqref{S1-L1}. We assume that it is true for arbitrary $L\in \mathbb{N}$ and we prove that it holds for $L+1$. 
		\begin{align*}
			\left(\mathcal{S}_{1}^{-1}\right)^{\otimes L}\otimes \mathcal{S}_{1}^{-1}&=
			\sum_{\tau_{1}=1}^{N}\ldots\sum_{\tau_{L}=1}^{N}\left\{e_{\tau_{1},\ldots,\tau_{L}}^{\tau_{1},\ldots,\tau_{L}}\right. \\&+\left. \sum_{b=1}^{L}(-1)^{b}\sum_{q_{1}=1}^{L}\sum_{q_{2}=q_{1}+1}^{L}\ldots\sum_{q_{b}=q_{b-1}+1}^{L}\left(\prod_{r=1}^{b}\delta_{s_{q_{r}},N}\right)\sum_{d_{q_{1}}=1}^{N-1}\ldots\sum_{d_{q_{r}}=1}^{N-1}e_{\theta_{1},\ldots,\theta_{L}}^{\eta_{1},\ldots,\eta_{L}}\right\}
			\\&\otimes
			\left\{ \sum_{\tau_{1}=1}^{N}e_{\tau_{L+1}}^{\tau_{L+1}}+(-1)\delta_{\tau_{L+1},N}\sum_{d_{L+1}=1}^{N-1}e_{1}^{d_{L+1}}\right\}
			\\&=
			\sum_{\tau_{1}=1}^{N}\ldots\sum_{\tau_{L+1}=1}^{N}\left\{e_{\tau_{1},\ldots,\tau_{L+1}}^{\tau_{1},\ldots,\tau_{L+1}}\right. \\&+\left. \sum_{b=1}^{L+1}(-1)^{b}\sum_{q_{1}=1}^{L+1}\sum_{q_{2}=q_{1}+1}^{L+1}\ldots\sum_{q_{b}=q_{b-1}+1}^{L+1}\left(\prod_{r=1}^{b}\delta_{s_{q_{r}},N}\right)\sum_{d_{q_{1}}=1}^{N-1}\ldots\sum_{d_{q_{b}}=1}^{N-1}e_{\theta_{1},\ldots,\theta_{L+1}}^{\eta_{1},\ldots,\eta_{L+1}}\right\}
		\end{align*}
		\begin{comment}

@@

indeed, in each coordinate $q_{r}$ with $r=1,\ldots b$ the species $N$ is replaced by all other possible species from $1$ to $N-1$.  Finally, considering all possible $\tau_{1},\ldots,\tau_{L}\in \{1,\ldots,N\}$, we obtain \eqref{steadyStateH}. 
